---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r message=FALSE, warning=FALSE}
library(RSQLite)
library(tidyverse)
set.seed(42)
conn = dbConnect(SQLite(), "ncaab")
# Filter out those who didn't play much and change height units to inches, remove duplicate rows
ncaa = read.csv("data/NCAA.csv")

# Read in draft data scraped from BR-ref.com and only take players from 2006 or earlier
draft = read.csv("data/Draft.csv") %>% rename("name" = "Player","team" = "School.Country") %>% filter(Year >= 2006)
# Get all individual NBA stats from 2010 and beyond, only guards/smaller wings
nba = read.csv("data/NBA.csv") %>% filter(Year >= 2010, Pos %in% c("PG", "SG","SF"))

```

### Get only players who played in the NBA as well as this upcoming draft class

```{r}
prospects = c("Anthony Edwards", "Tyrese Haliburton","Kira Lewis Jr.", "Josh Green", "Cole Anthony", "Tyrese Maxey", "Tyrell Terry", "Grant Riller", "Nico Mannion", "Malachi Flynn", "Cassius Winston", "Aaron Nesmith", "Jahmius Ramsey","Devon Dotson","Elijah Hughes","Immanuel Quickley", "CJ Elleby", "Isaiah Joe", "Cassius Stanley", "Tre Jones", "Payton Pritchard", "Ashton Hagans","Nate Hinton", "Markus Howard", "Desmond Bane", "Robert Woodard", "Saben Lee") 
#prospects = ncaa %>% filter(season == "2019-20", position == "Guard", height <= 78, games_started >= 20) %>% select(name) %>% unlist(use.names = F)
```


```{r}
# Get the names of all current NBA players
nba$Player %>% unique() -> nba.players
# Get 2019 Class and add it to pool of players along with 2020 class
draft %>% filter(Year == 2019) %>% select(name) %>% unlist(use.names = F) %>% c(nba.players,prospects) -> nba.players
# Get college stats of chosen NBA players
intersect(nba.players, unique(ncaa$name)) -> use.players
ncaa %>% filter(name %in% use.players, height <= 78) -> player
player %>% group_by(name, team, position) %>% summarise_all(last) %>%  ungroup() -> player
```


# Missing Values

Looks like earlier years had some issues with advanced stats that we may use, therefore it may make sense to impute or drop them all together.

```{r}
library(naniar)
# Create a dataframe specifically to visualize NAs
player %>% filter(position=="Guard") -> guard.na
# Visualize the NAs in that dataframe
gg_miss_fct(x = guard.na, fct = year)
```



```{r}
# Get rid of those that do not have DBPM and visualize them and make sure the only position used in 'Guards'
guards = player %>% filter(position == "Guard", games_played >= 20) %>% drop_na(defensive_box_plus_minus)
gg_miss_var(guards) + labs(title = "Missing Values by Variable")
# Drop players without a height value
player = player %>% drop_na(height)
na.omit(player) -> player
```


```{r}
# Get data ready for correlation matrix by only taking numeric columns
player %>% filter(position == "Guard") %>% select_if(is.numeric) %>%
  select(-c(games_played,games_started,minutes_played)) -> x
```


```{r}
library(corrplot)
corrmat = cor(scale(x))
corrplot(corrmat,order = "hclust", type = "upper", sig.level =  0.01)
```

Okay, so this is clearly a lot. How can we reduce the number of dimensions in the dimensions to make things 


```{r}
corrmat["defensive_win_shares",] %>% sort(decreasing = T)
```



Let's break down where we can get rid of variables by what skills they show us





Get rid of most all-in-one and volume stats 


```{r}
# Drop columns associated with counting stats, or that don't account for guard playing style
rebs = player %>% select(contains("rebound")) %>% colnames()
drop.cols = c(rebs,'field_goal_attempts','points','win_shares','box_plus_minus','two_point_attempts','free_throw_attempts','field_goals','two_pointers','free_throws', 'personal_fouls',"total_rebound_percentage", 'offensive_win_shares','player_efficiency_rating','win_shares_per_40_minutes','true_shooting_percentage','field_goal_percentage','three_point_attempts','turnovers','steals',"three_pointers","offensive_box_plus_minus","assists","blocks","free_throw_percentage","block_percentage","defensive_win_shares","steal_percentage", "weight", "two_point_percentage","free_throw_attempt_rate","three_point_percentage","total_rebound_percentage", "true_shooting_percentage", "height","year", "turnover_percentage","ft","X","points_produced")
x %>% select(-drop.cols)-> new.x



# New Correlations
cor(new.x)
cor(new.x) %>% corrplot(order = "hclust", type = "upper", sig.level =  0.01)

```


Hierarchical Clustering


```{r}
# Function to normalize data
normalize <- function(x) {
return ((x - min(x)) / (max(x) - min(x)))
}

```



```{r}
# Calculate correlation-based distance
dd = as.dist(1-cor(t(scale(new.x))))
# Calculate "complete" linkage
cor.complete = hclust(dd, method="complete")
# Calculate "average" linkage
cor.average = hclust(dd, method = "average")
# Plot "average" linkage 
plot(cor.average, main = "Average: correlation-based distance", xlab = "", labels = F)
# Plot "complete" linkage
plot(cor.complete,main = "Complete: correlation-based distance",xlab = "",labels = F)
```




# Create 5 clusters from dendrogram
```{r}
avg.clusters = cutree(cor.average, k = 6)
```


```{r}
# Pick Guards only and assign clusters to data frame
guards = player %>% filter(position == "Guard") %>% mutate(avg.cluster = avg.clusters)
#Join Draft data to college stats
guards = left_join(guards,draft[,c("name","Pick","team")], by = c("name","team"))
# If the draft info for that player is unavailable, assume undrafted and label the pick# as 61
guards$Pick[is.na(guards$Pick)] <- 61
guards$Pick[guards$Pick > 61] <- 61
# If the player's last season was the most recent one, label as 0 - current draft prospect
guards$Pick[guards$year == 2020] <- 61
# Let's see the qualities of each group
guards %>% select(colnames(new.x),avg.cluster) %>% group_by(avg.cluster) %>% summarise_each(mean) -> group_means
group_means
```


```{r}
guards %>% filter(avg.cluster == 1) %>% select(name,year,team, Pick,colnames(new.x)) %>% group_by(year)%>% arrange(Pick)
```


```{r}
guards %>% filter(year == 2020) %>% select(name,year,team, avg.cluster, colnames(new.x)) %>% group_by(avg.cluster)
```




## Graphs and Viz


```{r}
# Normalize data
guards$avg.cluster = as.factor(guards$avg.cluster)
guards %>% mutate_if(is.numeric, normalize) -> norm.data
norm.data %>% select(colnames(new.x),avg.cluster) %>% group_by(avg.cluster) %>% summarise_each(mean)

# Normalize group means
group_means$avg.cluster = as.factor(group_means$avg.cluster) 
group_means %>% mutate_if(is.numeric, normalize) -> group.norm

group.norm
# Create Radar Charts in Excel
#write.csv(group.norm, file = "grouped_norm.csv")
```




```{r}
library(ggthemes)
## LOLLIPOP CHARTS for 1 

library(reshape2)
# Clean up column names a bit
colnames(group.norm) = colnames(group.norm) %>% gsub(pattern = "_",replacement = " ")
# Melt dataframe to make it suitable for graphing and then GRAPH
melt(group.norm,id.vars = 'avg.cluster') %>% arrange(value) -> melt.norm

# Melted 'mean' data
colnames(group_means) = colnames(group_means) %>% gsub(pattern = "_",replacement = " ")
group_means %>% melt(id.vars = "avg.cluster") %>% arrange(value) -> melt.mean

# Get normalized and unscaled data
inner_join(melt.norm, melt.mean, by = c("avg.cluster","variable")) -> melt.total
```


```{r}
melt.total %>% filter(avg.cluster == 1) %>% ggplot(aes(x = variable, y = value.x, label = round(value.y,digits = 2)))+
    geom_segment( aes(x=variable ,xend=variable, y=0, yend=value.x), color="grey") +
    geom_point(size=10, color="#69b3a2") +
    coord_flip() +
  geom_text(color = 'white', size = 3) +
    theme(
      panel.grid.minor.y = element_blank(),
      panel.grid.major.y = element_blank(),
      legend.position="none",
      axis.ticks.x=element_blank(),
      axis.text.x=element_blank()
    ) +
    ylim(0,1) +
    labs(title = "Low Efficiency 3-and-D Guards",caption = "", y = "",x = "", subtitle = "Group Averages") 
```




```{r}
melt.total %>% filter(avg.cluster == 2) %>% ggplot(aes(x = variable, y = value.x, label = round(value.y,digits = 2)))+
    geom_segment( aes(x=variable ,xend=variable, y=0, yend=value.x), color="grey") +
    geom_point(size=10, color="#69b3a2") +
    coord_flip() +
  geom_text(color = 'white', size = 3) +
    theme(
      panel.grid.minor.y = element_blank(),
      panel.grid.major.y = element_blank(),
      legend.position="none",
      axis.ticks.x=element_blank(),
      axis.text.x=element_blank()
    ) +
    ylim(0,1) +
    labs(title = "Floor Generals", y = "",x = "", subtitle = "Group Averages") 
```





```{r}
melt.total %>% filter(avg.cluster == 3) %>% ggplot(aes(x = variable, y = value.x, label = round(value.y,digits = 2)))+
    geom_segment( aes(x=variable ,xend=variable, y=0, yend=value.x), color="grey") +
    geom_point(size=10, color="#69b3a2") +
    coord_flip() +
  geom_text(color = 'white', size = 3) +
    theme(
      panel.grid.minor.y = element_blank(),
      panel.grid.major.y = element_blank(),
      legend.position="none",
      axis.ticks.x=element_blank(),
      axis.text.x=element_blank()
    ) +
    ylim(0,1) +
    labs(title = "High Volume Scorers",y = "",x = "", subtitle = "Group Averages") 
```




```{r}
melt.total %>% filter(avg.cluster == 4) %>% ggplot(aes(x = variable, y = value.x, label = round(value.y,digits = 2)))+
    geom_segment( aes(x=variable ,xend=variable, y=0, yend=value.x), color="grey") +
    geom_point(size=10, color="#69b3a2") +
    coord_flip() +
  geom_text(color = 'white', size = 3) +
    theme(
      panel.grid.minor.y = element_blank(),
      panel.grid.major.y = element_blank(),
      legend.position="none",
      axis.ticks.x=element_blank(),
      axis.text.x=element_blank()
    ) +
    ylim(0,1) +
    labs(title = "Offensive Focal Points", y = "",x = "", subtitle = "Group Averages") 
```





```{r}
melt.total %>% filter(avg.cluster == 5) %>% ggplot(aes(x = variable, y = value.x, label = round(value.y,digits = 2)))+
    geom_segment( aes(x=variable ,xend=variable, y=0, yend=value.x), color="grey") +
    geom_point(size=10, color="#69b3a2") +
    coord_flip() +
  geom_text(color = 'white', size = 3) +
    theme(
      panel.grid.minor.y = element_blank(),
      panel.grid.major.y = element_blank(),
      legend.position="none",
      axis.ticks.x=element_blank(),
      axis.text.x=element_blank()
    ) +
    ylim(0,1) +
    labs(title = "Primary Ballhandlers with Lockdown D", y = "",x = "", subtitle = "Group Averages") 
```






```{r}
melt.total %>% filter(avg.cluster == 6) %>% ggplot(aes(x = variable, y = value.x, label = round(value.y,digits = 2)))+
    geom_segment( aes(x=variable ,xend=variable, y=0, yend=value.x), color="grey") +
    geom_point(size=10, color="#69b3a2") +
    coord_flip() +
  geom_text(color = 'white', size = 3) +
    theme(
      panel.grid.minor.y = element_blank(),
      panel.grid.major.y = element_blank(),
      legend.position="none",
      axis.ticks.x=element_blank(),
      axis.text.x=element_blank()
    ) +
    ylim(0,1) +
    labs(title = "Efficient 3 and D Wings", y = "",x = "", subtitle = "Group Averages") 
```






```{r}
library(reshape2)
# Clean up column names a bit
colnames(group.norm) = colnames(group.norm) %>% gsub(pattern = "_",replacement = " ")
# Melt dataframe to make it suitable for graphing and then GRAPH
melt.total %>% ggplot(aes(x = variable, y = value.x))+ 
    geom_segment( aes(x=variable ,xend=variable, y=0, yend=value.x), color="grey") +
    geom_point(size=5, color="#69b3a2") +
    facet_wrap(~avg.cluster) +
    coord_flip() +
    theme(
      panel.grid.minor.y = element_blank(),
      panel.grid.major.y = element_blank(),
      legend.position="none",
      axis.ticks.x=element_blank(),
      axis.text.x=element_blank()
    ) +
    ylim(0,1) +
    labs(title = "Characteristics of Each Cluster",caption = "Scores for each variable are relative", y = "",x = "") 
    
```





# Individuals 

```{r}
norm.data %>% filter(name == "Tyrese Haliburton") %>% select(names(new.x)) %>% melt() %>% 
  ggplot(aes(x = variable, y = value))+
    geom_segment( aes(x=variable ,xend=variable, y=0, yend=value), color="grey") +
    geom_point(size=10, color="#69b3a2") +
    coord_flip() +
    theme(
      panel.grid.minor.y = element_blank(),
      panel.grid.major.y = element_blank(),
      legend.position="none",
      axis.ticks.x=element_blank(),
      axis.text.x=element_blank()
    ) +
    ylim(0,1) +
    labs(title = "Tyrese Haliburton", y = "",x = "") 
    
```






```{r}
norm.data %>% filter(name == "Anthony Edwards") %>% select(names(new.x)) %>% melt() %>% 
  ggplot(aes(x = variable, y = value))+
    geom_segment( aes(x=variable ,xend=variable, y=0, yend=value), color="grey") +
    geom_point(size=10, color="#69b3a2") +
    coord_flip() +
    theme(
      panel.grid.minor.y = element_blank(),
      panel.grid.major.y = element_blank(),
      legend.position="none",
      axis.ticks.x=element_blank(),
      axis.text.x=element_blank()
    ) +
    ylim(0,1) +
    labs(title = "Anthony Edwards", y = "",x = "") 
    
```






